---
title: "Predictive Modeling"
author: "Lucas Fox, Nathalie Kirsch, Vanaaisha Pamnani, and Cuong Pham Vu"
date: "5/6/2022"
output: html_document
urlcolor: blue
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r include = FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

### PPOL 670 | Final Project

This assignment is stored on the following [Github repository](https://github.com/nakirsch/PPOL670_FinalProject). This markdown file includes code for predictive modeling and principal components analysis.

```{r}
library(tidyverse)
library(tidycensus)
library(tidymodels)
library(ggplot2)
library(patchwork)
library(httr)
library(jsonlite)
library(sf)
library(janitor)
library(lubridate)
library(stringr)
library(factoextra)


theme_set(theme_minimal())
```


``` {r}

crime_merged_agg <- read_csv("crime_merged_agg.csv", show_col_types = FALSE) %>%
  select(-"...1")

acs_merged <- read_csv("acs_merged.csv", show_col_types = FALSE) %>%
  select(-"...1")

crime_acs_merged <- left_join(crime_merged_agg %>% filter(year < 2020), 
                              acs_merged %>% filter(year < 2020), 
                              by = c("tract","year"))

crime_acs_numeric <- crime_acs_merged %>% 
  select_if(is.numeric) %>%
  select(-contains(c("PM","M","PE"))) %>%
  ungroup()

crime_acs_numeric <- crime_acs_numeric %>%
  select_if(~ !any(is.na(.)))

```



```{r}

# LASSO MODEL PREDICTING CRIME COUNT WITH ALL PREDICTORS

set.seed(20220503)

crime_split <- initial_split(data = crime_acs_numeric %>% select(-year, - GEOID), prop = 0.8)

crime_train <- training(x = crime_split)

crime_test <- testing(x = crime_split)

crime_rec <- recipe(count ~ ., data = crime_train) %>%
  step_scale(all_predictors()) %>%
  step_center(all_predictors())

folds <- vfold_cv(data = crime_train, v = 10, repeats = 10)

lasso_grid <- grid_regular(penalty(), levels = 10)

lasso_mod <- linear_reg(
  penalty = tune(),
  mixture = 1) %>%
  set_engine("glmnet", path_values = lasso_grid$penalty)

lasso_wf <- workflow() %>%
  add_recipe(crime_rec)%>%
  add_model(lasso_mod)

get_glmnet_coefs <- function(x) {
  
  x %>% 
    extract_fit_engine() %>% 
    tidy(return_zeros = TRUE) %>% 
    rename(penalty = lambda)
  
}

parsnip_ctrl <- control_grid(extract = get_glmnet_coefs)

lasso_cv <- lasso_wf %>%
  tune_grid(
    resamples = folds,
    grid = lasso_grid,
    control = parsnip_ctrl
  )

bind_rows(
  LASSO = show_best(lasso_cv, n = 1),
  .id = "model"
)

lasso_coefs <- 
  lasso_cv %>% 
  select(id, .extracts) %>% 
  unnest(.extracts) %>% 
  select(id, .extracts) %>% 
  unnest(.extracts) %>%
  group_by(id, term, penalty) %>%
  slice(1) %>%
  ungroup()


lasso_best <- lasso_cv %>% select_best("rmse")

lasso_final <- finalize_workflow(
  lasso_wf,
  parameters = lasso_best
)

lasso_fit <- lasso_final %>%
  fit(data = crime_train)

#Calculating RMSE on testing data

predictions <- bind_cols(
    crime_test,
    predict(object = lasso_fit, new_data = crime_test)
)


rmse(data = predictions, truth = count, estimate = .pred)

predictions <- bind_cols(
    crime_acs_numeric,
    predict(object = lasso_fit, new_data = crime_acs_numeric)
)

#Caluclating RMSE on full dataset

rmse(data = predictions, truth = count, estimate = .pred)

# Predictions by year

predictions %>%
  group_by(year) %>% 
  summarize(actual = sum(count),
            predicted = sum(.pred))

tidy(lasso_fit) %>%
  filter(estimate != 0)

```

Our model has an RMSE of 110 on the testing data and an RMSE of 98.6 crimes on the full dataset


```{r}

# LASSO MODEL PREDICTING CRIME COUNT WITH SELECTED PREDICTORS

crimes_acs_main  <- crime_acs_merged %>%
  select(DP02_0059PE, DP02_0060PE, DP02_0061PE, DP02_0062PE, DP02_0063PE, DP02_0064PE, DP02_0065PE, DP02_0066PE, DP02_0067PE, DP02_0068PE, DP05_0037PE, DP05_0038PE, DP05_0039PE, DP05_0044PE, DP05_0052PE, DP05_0058PE, DP03_0086E, DP02_0016E, DP02_0017E, DP02_0002PE, DP02_0010PE, year, count, population)

crimes_acs_2019 <- crime_acs_merged %>%
  filter(year == 2019) %>%
  select(DP02_0059PE, DP02_0060PE, DP02_0061PE, DP02_0062PE, DP02_0063PE, DP02_0064PE, DP02_0065PE, DP02_0066PE, DP02_0067PE, DP02_0068PE, DP05_0037PE, DP05_0038PE, DP05_0039PE, DP05_0044PE, DP05_0052PE, DP05_0058PE, DP03_0086E, DP02_0016E, DP02_0017E, DP02_0002PE, DP02_0010PE, year, count, population)

crimes_acs_2020 <- crime_acs_merged %>%
  filter(year == 2020) %>%
  select(DP02_0059PE, DP02_0060PE, DP02_0061PE, DP02_0062PE, DP02_0063PE, DP02_0064PE, DP02_0065PE, DP02_0066PE, DP02_0067PE, DP02_0068PE, DP05_0037PE, DP05_0038PE, DP05_0039PE, DP05_0044PE, DP05_0052PE, DP05_0058PE, DP03_0086E, DP02_0016E, DP02_0017E, DP02_0002PE, DP02_0010PE, year, count, population)


crimes_acs_main <- crimes_acs_main %>%
  select_if(is.numeric) %>%
  ungroup() %>%
  select_if(~ !any(is.na(.)))
 
crimes_acs_2020 <- crimes_acs_2020 %>%
  select_if(is.numeric) %>%
  ungroup()

crimes_acs_2019 <- crimes_acs_2019 %>%
  select_if(is.numeric) %>%
  ungroup()

crimemain_split <- initial_split(data = crimes_acs_main  %>% select(-year), prop = 0.8)

crimemain_train <- training(x = crimemain_split)

crimemain_test <- testing(x = crimemain_split)

crimemain_rec <- recipe(count ~ ., data = crimemain_train) %>%
  step_scale(all_predictors()) %>%
  step_center(all_predictors())

folds1 <- vfold_cv(data = crimemain_train, v = 10, repeats = 5)

lassomain_grid <- grid_regular(penalty(), levels = 10)

lassomain_mod <- linear_reg(
  penalty = tune(),
  mixture = 1) %>%
  set_engine("glmnet", path_values = lasso_grid$penalty)

lassomain_wf <- workflow() %>%
  add_recipe(crimemain_rec)%>%
  add_model(lassomain_mod)

lassomain_cv <- lassomain_wf %>%
  tune_grid(
    resamples = folds1,
    grid = lasso_grid,
    control = parsnip_ctrl
  )

bind_rows(
  LASSO = show_best(lassomain_cv, n = 1),
  .id = "model"
)

lassomain_coefs <-
  lassomain_cv %>%
  select(id, .extracts) %>%
  unnest(.extracts) %>%
  select(id, .extracts) %>%
  unnest(.extracts) %>%
  group_by(id, term, penalty) %>%
  slice(1) %>%
  ungroup()

lassomain_best <- lassomain_cv %>% select_best("rmse")

lassomain_final <- finalize_workflow(
  lassomain_wf,
  parameters = lassomain_best
)

lassomain_fit <- lassomain_final %>%
  fit(data = crimemain_train)

predictions_main <- bind_cols(
    crimemain_test,
    predict(object = lassomain_fit, new_data = crimemain_test)
)

# Calculating RMSE for testing data

rmse(data = predictions, truth = count, estimate = .pred)

predictions19_main <- bind_cols(
    crimes_acs_main %>% select_if(~ !any(is.na(.))),
    predict(object = lassomain_fit, new_data = crimes_acs_main %>% select_if(~ !any(is.na(.)))
            )
)

# Calculating RMSE for full dataset

rmse(data = predictions19_main, truth = count, estimate = .pred)

# Predictions by year

predictions19_main %>%
  group_by(year) %>%
  summarize(actual = sum(count),
            predicted = sum(.pred))

tidy(lassomain_fit) %>%
  filter(estimate != 0)
```

The only variables with explanatory power are percent of population 25 years or older with a bachelor's degree, percent of population that identifies having two or more races, and tract population in 2010. In total, the model with selected variables has an RMSE of 110 on the testing data and an RMSE of 149 on the full dataset.

```{r}

#PCA ANALYSIS

crime_pca <- prcomp(crime_acs_numeric %>% select(-year, -count, -GEOID))

pca_data <- as_tibble(crime_pca$x) %>%
  select(PC1, PC2)

tract_pca <- bind_cols(
  crime_acs_merged, 
  select(pca_data, PC1, PC2)
)

ggplot() +
  geom_point(
    data = tract_pca, 
    mapping = aes(x = PC1, y = PC2, color = population),
    alpha = 0.5
  ) +
  facet_wrap(~year) +
  theme_minimal() +
  labs(
    title = "PC1 And PC2 Capture Clear Relationships Across Years",
  )
```